---
title: "Application of PTE methods to a Hypothetical Vaccine Trial"
author: "Florian Stijven"
date: '`r Sys.Date()`'
output: 
  bookdown::pdf_document2:
  bookdown::github_document2:
---


# Introduction

In this document, the analysis of a set of hypothetical vaccine trials is
described. This document supplements the review paper of the proportion of treatment
effect explained (PTE) as a measure of surrogacy. All data in this document are
hypothetical, but meant to be realistic. The goal of this document is **not** to
show how different versions of the PTE are estimated in practice. Instead, the
goal is to illustrate three conceptual points related to the PTE:

1. The PTE can be interpreted in a variety of ways. However, for the PTE to have
a relevant interpretation, unverifiable assumptions are **always** needed. 
2. The PTE is often defined independent of baseline covariates. Without taking
such baseline covariates into account, the PTE can often not be interpreted in
the desired manner.
3. The validity of the non-causal interpretation of the PTE is inherently linked
to the potential application trials. The validity of the causal interpretation 
does not involve assumptions regarding potential application trials.

In this hypothetical scenario, we have data from a single evaluation trial in
which the efficacy of a vaccine is evaluated. In this trial, the occurrence of
infection in the year after vaccination is the primary endpoint, i.e., the
true endpoint, $T$. At the same time, the neutralizing antibody level 2 weeks after
vaccination has been measured. This serves as a potential surrogate endpoint,
$S$.

Independent of the evaluation trial, we consider two potential application
trials. These are trials in which the results of the previous surrogacy analysis
are used to replace the true endpoint with the surrogate endpoint. The
population in these two application trials differs from the evaluation trial's
population in terms of age and general health at baseline. In addition, the
vaccine in the second application trial has a different mechanism of action.
These differences are meant to emulate a real-life setting where, indeed, there
may be important differences between trials for the same disease. This also puts
the spotlight on the reasoning required for justifying the use of a potential
surrogate endpoint in a new trial, and the potential pitfalls.

The remainder of the document is structured as follows. First, the data
generating model underlying the evaluation trial is explained. For this trial,
two versions of the PTE are then estimated and interpreted in a causal framework.
Next, the data generating model underlying the two potential application trials
is explained. The results of the surrogacy evaluation in the evaluation trial
are then used to predict the treatment effects in these two application trials
*using only surrogacy information*. The accuracy of these predictions is then
related to (violations of) assumptions and the PTE's that were estimated in the
evaluation trial.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```


```{r}
# The tidyverse packages are loaded for data manipulation and plotting.
library(tidyverse)
# The mediation package is loaded for computing the PTE and related quantities. 
library(mediation)
# Set a seed for reproducibility.
set.seed(1)
```

# Evaluation Trial

## Data Generating Model

The data generating model is based on the selection diagram in Figure
\@ref(fig:selection-diagram). The nodes in that diagram correspond to the
following variables:

* **Z**. This is the treatment indicator where $Z = 0$ corresponds to the control 
vaccine and $Z = 1$ corresponds to the experimental vaccine. Treatment is randomized
in this scenario. Therefore, there are no arrows into $Z$.
* **S**. This is the surrogate endpoint, i.e., the neutralizing antibody level.
Treatment has a causal effect on $S$, and $S$ has a causal effect on the true
endpoint. Therefore, the neutralizing antibody level is a mediator for the
effect of vaccination.
* **T**. This is the true endpoint where $T = 0$ corresponds to infection in the year following vaccination and $T = 1$ to no infection in this period.
* **X**. This represents the confounding variables for the causal effect of the
neutralizing antibody level on the infection outcome. In this data generating model, 
there are only 2 confounding variables. These variables are measured at baseline. 
  * Age, $X_1$. Older patients tend to have a weaker immune response to vaccination, and
  are more susceptible to infection.
  * General health at baseline, $X_2$. Patients in better general health tend to have a 
  stronger immune response to vaccination, and are less susceptible to infection.
* **A**. This is the trial indicator where $A = 1$ corresponds to the evaluation
trial and $A = 0$ corresponds to the application trial. 

```{r selection-diagram, echo=FALSE, fig.cap="Selection diagram underlying the data generating model. The evaluation and first application trial are consistent with this diagram, but the second application trial not.", out.width='50%'}
coords = list(
  x = c(Z = 0, S = 0.5, `T` = 1, X = 1.25, A = 0.75),
  y = c(Z = 0, S = 0.75, `T` = 0, X = 0.75, A = 1.5)
)
vaccine_trial_dag = ggdag::dagify(
  S ~ Z,
  S ~ X,
  S ~ A,
  `T` ~ S,
  `T` ~ X,
  `T` ~ Z,
  X ~ A,
  coords = coords
)
ggdag::ggdag(vaccine_trial_dag) +
  theme_void()
```

We now discuss the distributions of these variables in the evaluation trial.
Concomitant with these distributions, we provide the code that is used to
generate the data in the evaluation trial.

### Treatment 

The treatment is randomized in a 1:1 fashion such that there are 10.000 patients 
in each treatment arm.

```{r}
# Sample size per treatment arm is fixed.
n = 1e4
# Construct integer vector where the first n elements are zero, and the next n
# elements 1.
Z = c(rep(0L, n), rep(1L, n))
```

### Age and General Health

As mentioned previously, age and general health are baseline covariates that 
confound the causal effect of $S$ on $T$. These variables are distributed in the
evaluation trial as follows.

* Age, $X_1 | A = 1 \sim N(35, 7^2)$. This population is thus quite young (in 
comparison with the application trials, see further). 
* General health, $X_2 | A = 1 \sim N(5, 5^2)$. This population has a good general health
(in comparison with the application trials, see further).

Because of randomization, these two variables are independent of the treatment
variable. However, the distribution of these variables changes in the application
trials.

```{r}
# Generate observations for the baseline covariates.
age = rnorm(n = 2 * n, mean = 35, sd = 7)
health = rnorm(n = 2 * n, mean = 5, sd = 5)

# Combine the treatment vector and the baseline covariates into a tibble. 
eval_trial_tbl = tibble(
  Z = Z, 
  age = age,
  health = health
)
```

### Neutralizing Antibody Levels, $S$

As mentioned before, the vaccine in the evaluation trial mainly exerts its effect
through inducing an antibody response. In other words, the vaccine reduces the
risk of infection by increasing the patient's neutralizing antibody levels. 
In the evaluation trial, $S | Z, X, A = 1$ is a normal distribution with unit variance
and mean $E(S | Z, X, A = 1) = 1.5 \cdot Z - 0.05 \cdot (X_1 - 45) + 0.15 \cdot X_2$.

```{r}
# The surrogate endpoint is simulated taking into account the baseline
# covariates and the treatment assignment.
eval_trial_tbl = eval_trial_tbl %>%
  mutate(S = 1.5 * Z - 0.05 * (age - 45) + 0.15 * health + rnorm(n = 2 * n))
```

### Infection Status, $T$

The true endpoint is a binary endpoint that depends on the treatment, the antibody
levels, and the baseline covariates. Its distribution is completely determined by
the mean, which depends on the mentioned variables in the following way,

$$E(T|Z, S, U, A = 1) = expit(1 + 0.1 \cdot Z + 0.5 \cdot S - 0.03 \cdot (X_1 - 45) + 0.1 \cdot X_2).$$ 
```{r}
# The true endpoint is simulated in two steps. First, the mean as function of the
# covariates is computed. Given this conditional mean, the true endpoint is sampled
# from a Bernoulli distribution.
eval_trial_tbl = eval_trial_tbl %>%
  mutate(
    eta = 0.8 + 0.1 * Z + 0.5 * S - 0.03 * (age - 45) + 0.1 * health,
    infection_free = rbinom(
      n = 2 * n,
      size = 1,
      prob = 1 / (1 + exp(-1 * eta))
    )
  )
```

```{r}
# Add more informative variable names and labels.
eval_trial_tbl = eval_trial_tbl %>%
  mutate(Treatment = factor(
    Z,
    levels = 0:1,
    labels = c("Control", "Experimental")
  ))
```



## Surrogacy Analysis 

In this subsection, we first do several simple analyses of the evaluation trial
that will help us to interpret the results regarding the PTE. Key concepts
regarding the PTE in the review paper are also repeated for completeness. Next,
two versions of the PTE are estimated for these data: (i) the "usual" PTE that
ignores baseline covariates, $PTE_{WT}$, and (ii) the PTE that does take into
account baseline covariates, $PTE_{WT}^X$. These estimates are then interpreted
in the causal frameworks discussed in the paper.

### Preliminaries

We start by computing the marginal treatment effect. 

```{r}
# Proportion infection-free after 1 year in the control group.
eval_trial_prop0 = mean(eval_trial_tbl$infection_free[eval_trial_tbl$Z == 0])
# Proportion infection-free after 1 year in the experimental group.
eval_trial_prop1 = mean(eval_trial_tbl$infection_free[eval_trial_tbl$Z == 1])
```


The proportions infection-free
in the control and experimental groups are, respectively, `r round(eval_trial_prop0, 3)` and 
`r round(eval_trial_prop1, 3)`. The risk difference is thus `r round(eval_trial_prop1 - eval_trial_prop0, 3)`.
In this context, the treatment effect is often summarized in the vaccine efficacy (VE),
$VE = 1 - RR$ where $RR$ is the relative risk of infection in the experimental vs
control group. The VE in this trial is `r round( 1 - (1 - eval_trial_prop1) / (1 - eval_trial_prop0), 3)`.

In what follows, $PTE_{WT}$ is defined as $\frac{\Delta - \Delta_S}{\Delta}$
where 
$$\Delta = E(T_1 - T_0 | A = 1) $$ 
and
\begin{equation} 
\Delta_S = \int E(T_1 | S_1 = s, A = 1) \, d F_{S_0 | A}(s | a = 1) - \int E(T_0 | S_0 = s, A = 1) \, d F_{S_0 | A}(s | a = 1).
  (\#eq:residual-treatment-effect)
\end{equation} 
The above two quantities are termed the marginal (or total) treatment effect,
and the residual treatment effect, respectively. In light of the general
definition used throughout the paper, we thus have that $h(u) = u$ and
$g(\cdot)$ is the expectation.

From the definition of $PTE_{WT}$, it follows that $PTE_{WT}$ is close to 1 when
the residual treatment effect, $\Delta_S$, is close to 0. From the definition of
$\Delta_S$ in \@ref(eq:residual-treatment-effect) follows that $\Delta_S$ is
close to 0 when the regression functions of $T$ on $S$ are similar in both
treatment groups. Indeed, we can rewrite $\Delta_S$ as the weighted difference
of these two regression functions,
$$\Delta_S = \int E(T_1 | S_1 = s, A = 1) - E(T_0 | S_0 = s, A = 1) \, d F_{S_0 | A}(s | a = 1),$$
where the weights depend on the distribution of the surrogate in the control group.
These two regression functions are plotted next in Figure \@ref(fig:regression-functions-smooth).

```{r regression-functions-smooth, fig.cap="Regression function of the regression of the true endpoint on the surrogate endpoint in each treatment group separately. These regression functions are estimated by local regression. The corresponding density estimates of the distribution of the surrogate endpoint are superimposed on this plot.", out.width='70%'}
eval_trial_tbl %>%
  ggplot() +
  geom_smooth(aes(x = S, y = infection_free, color = Treatment), se = FALSE) + 
  geom_density(aes(x = S, color = Treatment)) +
  xlim(c(-2.5, 7.5)) + 
  theme_bw()
```

These regression functions differ between the treatment groups. Indeed,
the regression function in the control group is larger than in the experimental
group for all values of $S$. Hence, $PTE_{WT}$ will be larger than 1.

We can also include baseline covariates in the definition of the proportion explained.
This leads to $PTE_{WT}^X = \frac{\Delta - E\left\{\Delta_S(X)\right\}}{\Delta}$ where
$$\Delta_S(x) = \int E(T_1 | S_1 = s, X = x, A = 1) - E(T_0 | S_0 = s, X = x,  A = 1) \, d F_{S_0 | A}(s | a = 1).$$
When $X$ contains all confounders, then $E\left\{ \Delta_S(X) \right\}$
corresponds to the interventional direct effect and
$\Delta - E\left\{ \Delta_S(X) \right\}$ to the interventional indirect effect. Under the additional
cross-world independence assumption, these effects correspond to their natural
counterparts. The ratio of the interventional/natural indirect effect and the
total effect is classically termed the proportion mediated. As for the PTE, this
is actually is misnomer because it is not a proportion.

### Estimation of PTE

As explained in Appendix E of the paper, $PTE_{WT}$ can be estimated with
existing software for causal mediation analysis. In this document, we use the
`mediate()` function from the `mediation` package for this purpose. For a
conventional mediation analysis, this function relies on two regression models.

1. A regression model for the mediator given the treatment and the confounders
measured at baseline.
2. A regression model for the true endpoint given the treatment, the mediator
and the confounders measured at baseline.

For estimating the $PTE_{WT}$, we replace the mediator with the surrogate, and
we do not include baseline covariates. For estimating the $PTE_{WT}^X$, we do
include baseline covariates.

The corresponding models are fitted in turn next. For the surrogate, we choose a
linear regression model. For the true endpoint, we choose a logistic regression
model. When we include the baseline covariates in these models, they are
correctly specified. However, the logistic regression model is misspecified when
we do not include the baseline covariates.

```{r}
# Fit the models that do not include the baseline covariates.
surrogate_marg_model = lm(S ~ Z, data = eval_trial_tbl)
true_endpoint_marg_model = glm(infection_free ~ Z + S, data = eval_trial_tbl,
                               family = binomial())
# Fit the models that include the baseline covariates.
surrogate_model = lm(S ~ Z + age + health, data = eval_trial_tbl)
true_endpoint_model = glm(infection_free ~ Z + S + age + health,
                          data = eval_trial_tbl,
                          family = binomial())
```

We can now estimate $PTE_{WT}$ and $PTE_{WT}^X$. This is done in the two next
code chunks. For completeness, we call the summary method on the object returned
by `mediate()`. This method returns numerous estimates, which are discussed in
turn. We only focus on the estimates with `(treated)` in their name. These
estimates correspond to components of $PTE_{WT}$ as defined previously. If we
switch both treatment arms, then the estimates with `(control)` in their name
are obtained.

* Average causal mediation effect, `ACME (treated)`. This is often termed the indirect effect.
The estimator in
`mediate()` is only consistent for the indirect effect if all confounders are
included in the models in `mediate()`. 
  * This estimator is in any case consistent for $\Delta - \Delta_S$ in the 
  definition of $PTE_{WT}$ when no baseline covariates have been included and the models 
  are correctly specified.
  * This estimator is in any case consistent for $\Delta - E\left\{\Delta_S(U) \right\}$ 
  in the definition of $PTE_{WT}^X$ when baseline covariates have been included and the models 
  are correctly specified.
* Average direct effect, `ADE (treated)`. As for the indirect effect, the estimator in `mediate()` is
only consistent for the direct effect if all confounders have been included in the
models in `mediate()`. 
  * This estimator is in any case consistent for $\Delta_S$ in the 
  definition of $PTE_{WT}$ when no baseline covariates have been included and the models 
  are correctly specified.
  * This estimator is in any case consistent for $E\left\{\Delta_S(U) \right\}$ 
  in the definition of $PTE_{WT}^X$ when baseline covariates have been included and the models 
  are correctly specified.
* `Total effect`. This estimator is in any case consistent for $\Delta$ because
treatment is randomized.
* Proportion mediated, `Prop. Mediated (treated)`. This estimator is the ratio
of `ACME (treated)` and `Total effect`. 
  * This is a consistent estimator for $PTE_{WT}$ when no baseline covariates 
  have been included and the models 
  are correctly specified.
  * This is a consistent estimator for $PTE_{WT}^X$ when baseline covariates 
  have been included and the models 
  are correctly specified.

```{r}
PTE_WT = mediate(
  # Model for the mediator, i.e., the surrogate.
  model.m = surrogate_marg_model,
  # Model for the outcome, i.e., the true endpoint.
  model.y = true_endpoint_marg_model,
  # Number of MC samples for the numerical approximation.
  sims = 1e3,
  # Name of the treatment variable.
  treat = "Z",
  # Name of the mediator (surrogate) variable.
  mediator = "S"
)
summary(PTE_WT)
```

```{r}
PTE_WT_X = mediate(
  model.m = surrogate_model,
  model.y = true_endpoint_model,
  sims = 1e3,
  treat = "Z",
  mediator = "S"
)
summary(PTE_WT_X)
```

### Interpretation of Estimates

The relevant results are summarized in the Table \@ref(tab:results-pte). We
discuss these estimates in turn. Sampling variability is ignored.

* The estimated marginal treatment effect, $\Delta$, quantified by the risk difference, is
$0.069$ or $0.071$. The estimate changes slightly when ignoring versus including
baseline covariates. This provides the reference to compare the other quantities
with.
* The residual treatment effect, $\Delta_S$, is $-0.015$. Under the conditional 
independence assumption, $S_0 \perp T_1 | S_1$ and $S_1 \perp T_0 | S_0$, this
is equal to the *causal residual treatment effect*, $\Delta_{S, c}$. If $S$ is a 
principal surrogate, then $\Delta_{S, c} = 0$ by Lemma 3.1 in the review paper. 
  * This quantity **cannot** be interpreted as a direct effect because there is 
  confounding for the causal effect of $S$ on $T$ by age and general health.
* The $PTE_{WT}$ is $1.47$. In this context, principal surrogacy may not be satisfied exactly. So,
$\Delta_{S, c}$ may be close to zero, but not exactly zero. The $PTE_{WT} =
\frac{\Delta - \Delta_{S, c}}{\Delta}$ provides a natural scale to assess
closeness to zero. Indeed, closeness to zero is quantified relative to the
marginal treatment effect.
  * Under the conditional independence assumption, $S$ is thus not very close
  to being a principal surrogate.
* The $E(\Delta_S(X))$ is $0.005$. Because $X$ contains all confounders, this
is a valid estimate of the interventional direct effect. If we additionally make
the cross-world independence assumption, this is also a valid estimate of the
natural direct effect.
  * The direct effect of treatment is small. Therefore, a large portion of the
  treatment effect is mediated by the surrogate.
* The $PTE_{WT}^X$ is $0.870$. Because we have controlled for all confounders in
the computation of this quantity, this is a valid estimate of the proportion
mediated. Most of the treatment effect is thus mediated by the surrogate
endpoint.

```{r results-pte, echo=FALSE}
results_pte =
  tibble(
    "$\\Delta_S$" = PTE_WT$z1,
    "$\\Delta$" = PTE_WT$tau.coef,
    "$PTE_{WT}$" = PTE_WT$n1,
  )
results_pte_x =
  tibble(
    "$E(\\Delta_S(X))$" = PTE_WT_X$z1,
    "$\\Delta$" = PTE_WT_X$tau.coef,
    "$PTE_{WT}^X$" = PTE_WT_X$n1,
  )
knitr::kable(
  list(
    results_pte,
    results_pte_x
  ),
  caption = "Estimates for the proportions explained and related quantities. The first and second table correspond to the analysis ignoring and including baseline covariates, respectively.",
  escape = FALSE,
  digits = 3, 
  booktabs = TRUE
)
```



# Application Trials

In this section, we introduce two potential application trials. First, the data
generating model for these trials is explained. Next, we show how the quantities
estimated in Table \@ref(tab:results-pte) can be interpreted in reference to
these application trials. Indeed, under a set of assumptions, these quantities
tell us how good we can predict the treatment effects in new trials using only
the surrogate endpoint and possibly baseline covariates.

```{r}
eval_trial_tbl  = eval_trial_tbl %>%
  mutate(S_tilde = predict(true_endpoint_model, newdata = eval_trial_tbl %>%
                             mutate(Z = 1), type = "response"))
eval_trial_tbl %>%
  group_by(Z) %>%
  summarize(predicted_proportion = mean(S_tilde), proportion = mean(infection_free)) %>%
  pivot_longer(cols = 2:3) %>%
  pivot_wider(names_from = "Z", values_from = "value") %>%
  mutate(Delta = `1` - `0`)
```

## Data Generating Model

In the following subsection, we discuss the differences between the evaluation
and application trials. The application trials have $10.000$ patients in each
treatment arm.

### Application Trial 1

The first application trial is consistent with the selection diagram in
Figure \@ref(fig:selection-diagram). This application trial thus only differs
from the evaluation trial in the distribution of the baseline covariates and the
distribution of $S | Z, X$:

* The population in application trial 1 is older and less healthy than in the 
evaluation trial.
  * Age, $X_1 | A = 0 \sim N(40, 7^2)$. 
  * General health, $X_2 | A = 0 \sim N(1, 5^2)$
* The control vaccine is assumed to be the same as in the evaluation trial. Therefore,
we have that $S | Z = 0, X = x, A = 0$ and $S | Z = 0, X = x, A = 1$ are equal
in distribution for all possible values of $x$. However, the experimental
vaccine now has a smaller effect on the neutralizing antibody level. Therefore,
$S | Z, X, A = 0$ is a normal distribution with unit variance and mean $E(S | Z,
X, A = 0) = 0.75 \cdot Z - 0.05 \cdot (X_1 - 45) + 0.15 \cdot X_2$. 

```{r}
# Generate observations for the baseline covariates. These distributions differ
# from those in the evaluation trial.
age = rnorm(n = 2 * n, mean = 40, sd = 5)
health = rnorm(n = 2 * n, mean = 2, sd = 4)

# Combine the treatment vector and the baseline covariates into a tibble. Note 
# that the treatment variables, Z, is the same as for the evaluation trial.
appl_trial1_tbl = tibble(
  Z = Z, 
  age = age,
  health = health
)
# The surrogate endpoint is simulated. the distribution of the surrogate as a
# function of treatment and baseline covariates is different from the evaluation
# trial.
appl_trial1_tbl = appl_trial1_tbl %>%
  mutate(S = 0.75 * Z - 0.05 * (age - 45) + 0.15 * health + rnorm(n = 2 * n))
# The true endpoint is simulated as in the evaluation trial.
appl_trial1_tbl = appl_trial1_tbl %>%
  mutate(
    eta = 0.8 + 0.1 * Z + 0.5 * S - 0.03 * (age - 45) + 0.1 * health,
    infection_free = rbinom(
      n = 2 * n,
      size = 1,
      prob = 1 / (1 + exp(-1 * eta))
    )
  )
# Add more informative variable names and labels.
appl_trial1_tbl = appl_trial1_tbl %>%
  mutate(Treatment = factor(
    Z,
    levels = 0:1,
    labels = c("Control", "Experimental")
  ))
```

### Application Trial 2

The second application trial is **not** consistent with the selection diagram in
Figure \@ref(fig:selection-diagram). This application trial differs
from the evaluation trial in the distribution of the baseline covariates, the
distribution of $S | Z, X$, and the distribution of $T | Z, X, S$. The latter
is not allowed by the selection diagram in Figure \@ref(fig:selection-diagram).

* The population in application trial 2 is the same as in application trial 1. So,
the distribution of baseline covariates is also the same.
* The control vaccine is assumed to be the same as in the evaluation trial. Therefore,
we have that $S | Z = 0, X = x, A = 0$ and $S | Z = 0, X = x, A = 1$ are equal
in distribution for all possible values of $x$. However, the experimental
vaccine now mainly operates through cellular immunity. This type of immunity has
not been measured. Therefore, $S | Z, X, A = 0$ is a normal distribution with
unit variance and mean 
$E(S | Z, X, A = 0) = 0.3 \cdot Z - 0.05 \cdot (X_1 - 45) + 0.15 \cdot X_2$. 
Indeed, the treatment effect on the surrogate is now much smaller.
* Since the experimental vaccine mainly operates through cellular immunity, which was 
not measured, the direct effect is now important. This is reflected in the distribution 
of $T | Z, X, S, A = 0$. Indeed, the corresponding mean is
$$E(T | Z, S, U, A = 0) = expit \left( 1 + 2.5 \cdot Z + 0.5 \cdot S - 0.03 \cdot (X_1 - 45) + 0.1 \cdot X_2 \right)$$

```{r}
# Generate observations for the baseline covariates. These distributions are the
# same as in application trial 1.
age = rnorm(n = 2 * n, mean = 40, sd = 5)
health = rnorm(n = 2 * n, mean = 2, sd = 4)

# Combine the treatment vector and the baseline covariates into a tibble. Note 
# that the treatment variables, Z, is the same as for the evaluation trial.
appl_trial2_tbl = tibble(
  Z = Z, 
  age = age,
  health = health
)
# The surrogate endpoint is simulated. the distribution of the surrogate as a
# function of treatment and baseline covariates is different from the evaluation
# trial.
appl_trial2_tbl = appl_trial2_tbl %>%
  mutate(S = 0.1 * Z - 0.05 * (age - 45) + 0.15 * health + rnorm(n = 2 * n))
# The true endpoint is simulated. This conditional distribution differs from the
# evaluation trial. This is not conssitent with the previoulsy mentioned
# selection diagram.
appl_trial2_tbl = appl_trial2_tbl %>%
  mutate(
    eta = 1 + 1 * Z + 0.5 * S - 0.03 * (age - 45) + 0.1 * health,
    infection_free = rbinom(
      n = 2 * n,
      size = 1,
      prob = 1 / (1 + exp(-1 * eta))
    )
  )
# Add more informative variable names and labels.
appl_trial2_tbl = appl_trial2_tbl %>%
  mutate(Treatment = factor(
    Z,
    levels = 0:1,
    labels = c("Control", "Experimental")
  ))
```



## Prediction of Treatment Effects

As explained in the review paper, $PTE_{WT}$ and $PTE_{WT}^X$ each imply a 
prediction rule $\mu$ and $\mu_X$ for the true endpoint. In the context of the
application trials, these are prediction rules for the true endpoint, given the
observed surrogate endpoint and possibly baseline covariates:
$$\mu(s) = E(T | Z = 1, S = s, A = 1) \text{ and } \mu_x(s) = E(T | Z = 1, S = s, X = x, A = 1)$$
This is a particular transformation of $(S, X)$, e.g., $\Tilde{S} = \mu_X (S)$. 
For $PTE_{WT}$ we drop the dependence on $X$.

Under a set of assumptions, $PTE_{WT}$ and $PTE_{WT}^X$ quantify the relative
discrepancy between the treatment effect on $\Tilde{S}$ and $T$. This is shown
in Lemma 4.2 in the review paper. In Table
\@ref(tab:application-trial-pte-results), the "true" and predicted treatment
effects are summarized in both application trials. The predicted treatment effects
are defined as follows,
$$\Tilde{\Delta}^* = E \left( \mu(S_1) - \mu(S_0) | A = 0 \right)$$
and
$$E\left( \Tilde{\Delta}^*(X)  \right) = E \left( \mu_X(S_1) - \mu_X(S_0) | A = 0 \right). $$

```{r}
# Combine application trials into a single data set.
appl_trials_tbl = 
  bind_rows(
    appl_trial1_tbl %>%
      mutate(Trial = 1L),
    appl_trial2_tbl %>%
      mutate(Trial = 2L)
  )
# Compute the predicted treatment effects in both application trials.
appl_trials_tbl = appl_trials_tbl %>%
  mutate(
    S_tilde = predict(
      # Logistic regression model without baseline covariate.
      true_endpoint_marg_model,
      newdata = pick(everything()) %>%
        mutate(Z = 1L), # The prediction rule from the evaluation trial is used.
      type = "response"
    ),
    S_tilde_X = predict(
      # Logistic regression model with baseline covariates.
      true_endpoint_model,
      newdata = pick(everything()) %>%
        mutate(Z = 1L), # The prediction rule from the evaluation trial is used.
      type = "response"
    )
  )
```

```{r application-trial-pte-results, echo=FALSE}
appl_trials_tbl %>%
  group_by(Treatment, Trial) %>%
  summarize(
    prop_infection_free = mean(infection_free),
    mean_S_tilde = mean(S_tilde),
    mean_S_tilde_X = mean(S_tilde_X)
  ) %>%
  pivot_wider(names_from = Treatment, values_from = 3:5) %>%
  mutate(
    "$\\Delta$" = prop_infection_free_Experimental - prop_infection_free_Control,
    "$\\tilde{\\Delta}^*$" = mean_S_tilde_Experimental - mean_S_tilde_Control,
    "$E \\left( \\tilde{\\Delta}^*(X) \\right)$" = mean_S_tilde_X_Experimental - mean_S_tilde_X_Control
  ) %>%
  dplyr::select(c(1, 8:10)) %>%
  knitr::kable(digits = 3,
               escape = FALSE,
               booktabs = TRUE,
               caption = "True and predicted treatment effects in the application trials. The mean difference is used as effect measure in this table.")
```


Alternatively, we can look at the results in terms of (predicted) VE. While Lemma
4.2 still holds for VE as effect measure for $PTE_{WT}$, issues arise for 
$PTE_{WT}^X$. Indeed, VE is a non-collapsible effect measure. So, the conditional 
residual effect $\Delta_S(x)$ and its expectation cannot be compared with the 
marginal treatment effect. In principle, we could redefine the total treatment effect
as $E(VE(X))$ for allowing for proper comparisons. However, it is unclear how such 
quantities can be properly interpreted. Therefore, the predicted treatment effects
are defined as follows,
$$ \Tilde{VE}^* = 1 - \frac{1 - E \left( \mu(S_1) | A = 0 \right)}{1 - E \left( \mu(S_0) | A = 0 \right)} $$
and 
$$ E\left( \Tilde{VE}^*(X)  \right) = 1 - \frac{1 - E \left( \mu_X(S_1) | A = 0 \right)}{1 - E \left( \mu_X(S_0) | A = 0 \right)}. $$

These results are summarized in Table \@ref(tab:application-trial-ve-results). 

```{r application-trial-ve-results, echo=FALSE}
appl_trials_tbl %>%
  group_by(Treatment, Trial) %>%
  summarize(
    prop_infection_free = mean(infection_free),
    mean_S_tilde = mean(S_tilde),
    mean_S_tilde_X = mean(S_tilde_X)
  ) %>%
  pivot_wider(names_from = Treatment, values_from = 3:5) %>%
  mutate(
    "$VE$" = 1 - (1 - prop_infection_free_Experimental) / (1 - prop_infection_free_Control),
    "$\\tilde{VE}^*$" = 1 - (1 - mean_S_tilde_Experimental) / (1 - mean_S_tilde_Control),
    "$E \\left( \\tilde{VE}^*(X) \\right)$" = 1 - (1 - mean_S_tilde_X_Experimental) / (1 - mean_S_tilde_X_Control)
  ) %>%
  dplyr::select(c(1, 8:10)) %>%
  knitr::kable(digits = 3,
               escape = FALSE,
               booktabs = TRUE,
               caption = "True and predicted treatment effects in the application trials. The vaccine efficacy is used as effect measure in this table.")
```


### Other Application Trials

The two application trials considered in this document have been cherry-picked
to illustrate potential pitfalls in using $PTE_{WT}$ and $PTE_{WT}^X$. These two
trials indeed represent extreme situations. In the first application trial, the
treatment operated through the same mechanism (neutralizing antibodies) as the
evaluation trial. The second application trial's treatment operated through a
very different mechanism, cellular immunity.

Reality is much more complex than presented in this document. Indeed, there
exist a plethora of different types of neutralizing antibodies. Additionally,
the same "type" of antibody can potentially be measured by different methods.
All these factors should be taken into account in the evaluation of the
assumptions of Lemma 4.2 in the review paper. Indeed, a $PTE_{WT}$ or
$PTE_{WT}^X$ close to 1 is *some* evidence of surrogacy, but the relevance of
this result depends on unverifiable extrapolating assumptions. Clearly, this is
a very difficult endeavor in which no consensus might be reached among
scientists. The only empirical way forward, is the extension to multiple trials
as touched upon in the Discussion of the review paper.

